---
title: "Getting started with R"
subtitle: "Session 3 of 4: Data visualisation"
author: "Freddie Heather"
format: 
  revealjs:
    eval: true
    echo: true
    fig.align: 'center'
project:
  output-dir: docs
---

# Recap from session 2

## Recap questions

::: incremental
-   What is a script?
-   Name five data manipulation functions from last week.
-   What is the 'pipe' operator?
-   What does "\>=", and "!=" do, and how would you use them?
-   I want to create a new column in a tibble. What function would I use?
-   I filter my data (e.g., age \>= 3), name three way to check if the filter has worked?
-   What is the 'tidyverse'?
:::

## Today: Session 3

-   Data import, checking, wrangling (recap)
-   Data visualisation
-   Critical thinking approach

::: callout-tip
## Task

1.  Start a new R project, or open the project from last week.
2.  Create a new script and call it something informative (e.g., `session3` or `data_vis`).
3.  Comment your name, date, and a title at the top of the script.
4.  **Save your script!**
:::

## R is a tool

```{css}
#| echo: false

.reveal table {
  font-size: smaller;
}

```

|   | Mechanics | Science |
|--------------------|------------------|----------------------------------|
| **Problem** | Bolt won't loosen by hand | I have hypothesis and data; I want results for a paper |
| **Solution** | Use a spanner/wrench/socket/hammer | Use functions in R to import, clean, visualise and model data |
| **Outcome** | Loosen bolt | Figures and statistics |

# Data import & checking

## Data import

::: callout-tip
## Task

1.  Download the Reef Life Survey file "cape_howe.csv" from [this link](https://github.com/FreddieJH/r_workshop/blob/main/cape_howe.csv)
2.  Import the csv and save it in R as an object (this will be your raw object and never overwritten)
3.  Create a new object that is only a single site - Choose one other than JBMP-S2
:::

```{r }

library(tidyverse)

capehowe <- 
  read_csv("cape_howe.csv")

jervis <- 
  capehowe %>% 
  filter(site_code == "JBMP-S2")

```

## Data import

-   Has it imported correctly?
-   Sometimes you might have your hypothesis before data collection, not always

::: callout-tip
## Task

1.  Use `head()` and at least one other function to check to see things look okay.
2.  How many columns? What do the columns mean?
3.  Think of a question(s) that you could answer using these data.
:::

# Data visualisation

## Two numerical variables

-   H1: I think that bigger fish are less abundant.
-   What columns would I need to consider?
-   How might I visualise this? boxplot? scatterplot? barplot?

## A bit of wrangling first

-   To answer the question if 'bigger' fish are more abundant, we are interested in the species, their mean size, and their total abundance.

```{r }

# getting the mean size and abundance of each species in jervis bay
sizes_byspp <- 
  jervis %>% 
  summarise(mean_size = weighted.mean(size_class, n_500m2), 
          total_abundance = sum(n_500m2), 
          .by = species_name) 

```

## Choosing the type of plot

|                              | Predictor variable (x) | Response variable (y) |
|--------------------------|------------------------|-----------------------|
| Barplot, Boxplot, Violinplot | Categorical            | Numeric               |
| Scatterplot                  | Numeric                | Numeric               |
| Density, Histogram           | Numeric                | \-                    |

## Scatterplot using `ggplot()`

-   `ggplot()` is a function from the `ggplot2` package (it's part of the `tidyverse` collection of packages)
-   Scatterplot is used to compare two numerical variables

## Mean size vs. total abundance

```{r }

sizes_byspp %>% 
  ggplot(aes(x = mean_size, y = total_abundance)) +
  geom_point()


# Its not clean, potentially just one outlier
# Would transformation of the data make it more obvious to visualise?
```

## Transforming data

```{r }
#| eval: false
# Two ways transform data

# transform before input into the ggplot
sizes_byspp %>% 
  mutate(log_tot_abun = log(total_abundance)) %>% 
  ggplot(aes(x = mean_size, y = log_tot_abun)) +
  geom_point()

# perform the transformation in the ggplot
sizes_byspp %>% 
  ggplot(aes(x = mean_size, y = total_abundance)) +
  geom_point() +
  scale_y_log10()

# what are the two differences between these two? (think: visual and data)

```

## Dealing with overalapping points

```{r }

# Changing the transparency of the point (= alpha)
sizes_byspp %>% 
  ggplot(aes(x = mean_size, y = total_abundance)) +
  geom_point(alpha = 0.5) + # alpha argument changes transparency
  scale_y_log10() 

# Changing the type of point (= pch)
sizes_byspp %>% 
  ggplot(aes(x = mean_size, y = total_abundance)) +
  geom_point(pch = 21) + # pch argument changes point type (Google: "pch in r")
  scale_y_log10() 

```

::: callout-tip
## Task

1.  Change alpha to various values
2.  Change pch to get a hollow square
:::

## A cleaner plot

```{r }

sizes_byspp %>% 
  ggplot(aes(x = mean_size, y = total_abundance)) +
  geom_point(pch = 21, alpha = 0.8) + 
  scale_y_log10() + 
  labs(x = "Species mean body size (cm)", 
       y = "Total observations")

# Would it be worth doing more stats?

```

## One numerical, one categorical

-   H2: More species are seen in Method 1, than Method 2 (inverts + cryptic species)
-   What columns do I need? What the classes of these columns?
-   What would be suitable plots for these data?

## Wrangling before plotting

```{r }

# say we only care about highly biodiverse sites (> 120)

# calculating the number of species at a site by the method
nspp_bysite <- 
  capehowe %>% 
  select(species_name, site_code, method) %>% 
  distinct() %>% 
  count(site_code, method) 

# number of species at each survey within a site for each method
nspp_bysurv_bysite <- 
  capehowe %>% 
  select(survey_date, species_name, site_code, method) %>% 
  distinct() %>% 
  count(survey_date, site_code, method) 

```

## Boxplot in R

```{r }

# a quick plot
nspp_bysurv_bysite %>% 
  ggplot() +
  aes(x = site_code, 
      y = n) +
  geom_boxplot()

```

## Too much data

```{r }

# say we only care about highly biodiverse sites (> 120 species)

selected_sites <- 
  nspp_bysite %>% 
  arrange(desc(n)) %>% 
  filter(n > 120) %>% 
  pull(site_code)

selected_sites
```

## ...and plot method 1 and 2 separately

```{r }
#| code-line-numbers: "7"

# separate by the method 
nspp_bysurv_bysite %>% 
  filter(site_code %in% selected_sites) %>% 
  ggplot() +
  aes(x = site_code, 
      y = n, 
      col = as.factor(method)) + # why do we need as.factor() here?
  geom_boxplot()

```

## Ordered boxplot in R

```{r }
#| code-line-numbers: "3"

nspp_bysurv_bysite %>% 
  filter(site_code %in% selected_sites) %>% 
  mutate(site_code = fct_reorder(site_code, n)) %>% 
  ggplot() +
  aes(x = site_code, 
      y = n, 
      col = as.factor(method)) +
  geom_boxplot()

```

## Violin plots

```{r }
#| code-line-numbers: "7"

nspp_bysurv_bysite %>% 
  filter(site_code %in% selected_sites) %>% 
  mutate(site_code = fct_reorder(site_code, n)) %>% 
  ggplot() +
  aes(x = site_code, 
      y = n) + 
  geom_violin() # how does this differ from a boxplot?

# Try and also separate the method by colour
```

## Barplots

```{r }
#| code-line-numbers: "11"

# For a barplot you might want the bar to represent the mean or median.
# how do barplots differ from violin or boxplots?
nspp_bysurv_bysite %>% 
  filter(site_code %in% selected_sites) %>% 
  summarise(mean_diversity = mean(n, na.rm = TRUE), 
            .by = c(site_code, method)) %>% 
  ggplot() +
  aes(x = site_code, 
      y = mean_diversity, 
      fill = as.factor(method)) +
  geom_col() +
  labs(fill = "Method", 
       x = "Site Code",
       y = "Mean number of species")

```

## One numerical variable

::::: columns
::: {.column width="50%"}
```{r }

nspp_bysite %>%
  ggplot() +
  aes(x = n, col = as.factor(method)) +
  geom_density()

```
:::

::: {.column width="50%"}
```{r }

nspp_bysite %>%
  ggplot() +
  aes(x = n, col = as.factor(method)) +
  geom_histogram(binwidth = 10)

```
:::
:::::

## Extra - multiple plots

-   Sometime you want to split up plots by a factor

```{r }

nspp_bysite %>%
  ggplot() +
  aes(x = n) +
  geom_histogram(binwidth = 10) +
  facet_wrap(~method)

```

## Extra - Error bars

```{r }
#| code-line-numbers: "5,12,13,14,15"

# Barplots need error bars 
nspp_bysurv_bysite %>% 
  filter(site_code %in% selected_sites) %>% 
  summarise(mean_diversity = mean(n, na.rm = TRUE), 
          sd_diversity = sd(n, na.rm = TRUE),
            .by = c(site_code, method)) %>% 
  ggplot() +
  aes(x = site_code, 
      y = mean_diversity, 
      fill = as.factor(method)) +
  geom_col(position = position_dodge(width = 1)) +
  geom_errorbar(aes(ymin = mean_diversity - sd_diversity, 
                    ymax = mean_diversity + sd_diversity), 
                 width = 0.2,
                position = position_dodge(width = 1)) +
  labs(fill = "Method", 
       x = "Site Code",
       y = "Mean number of species")
```

## Extra - Theming

```{r }
#| code-line-numbers: "19,20"

# Making pretty plots
nspp_bysurv_bysite %>% 
  filter(site_code %in% selected_sites) %>% 
  summarise(mean_diversity = mean(n, na.rm = TRUE), 
          sd_diversity = sd(n, na.rm = TRUE),
            .by = c(site_code, method)) %>% 
  ggplot() +
  aes(x = site_code, 
      y = mean_diversity, 
      fill = as.factor(method)) +
  geom_col(position = position_dodge(width = 1)) +
  geom_errorbar(aes(ymin = mean_diversity - sd_diversity, 
                    ymax = mean_diversity + sd_diversity), 
                 width = 0.2,
                position = position_dodge(width = 1)) +
  labs(fill = "Method", 
       x = "Site Code",
       y = "Mean number of species") +
  scale_fill_brewer(palette = "Set1") + #https://r-graph-gallery.com/38-rcolorbrewers-palettes.html
  theme_classic() # https://ggplot2.tidyverse.org/reference/ggtheme.html

```

## Extra - Layering geoms

```{r }
#| code-line-numbers: "9,10"

# Change the order of the geom_line() and geom_point()
nspp_bysurv_bysite %>% 
  mutate(year = year(survey_date)) %>%
  summarise(mean_diversity = mean(n, na.rm = TRUE), 
            .by = c(year, method)) %>% 
  ggplot() +
  aes(x = year, 
      y = mean_diversity) +
  geom_line(aes(group = as.factor(method))) +
  geom_point(aes(col = as.factor(method)), size = 5) 

```

## Cheatsheet

<https://statsandr.com/blog/files/ggplot2-cheatsheet.pdf>

## Next week

-   Collating all that we have learned.
-   Please bring your own data if you have some.